---
title: "speed_comparisons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{speed_comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  paged.print = FALSE
)
```

### Speed Comparisons

Below are some speed comparisons of various functions. More functions will get added to the speed comps over time.

A few notes:

* Comparing times from separate functions won't be very useful. For example - the `summarize()` tests were performed on a different dataset from `case_when()`.
* `setDTthreads(4)` was used for `data.table` & `tidytable` timings.
* Modify-by-reference was used in `data.table` when being compared to `mutate.()` & `dplyr::mutate()`
* `fill.()` & `tidyr::fill()` both work with character/factor/logical columns, whereas `data.table::nafill()` does not. Testing only included numeric columns due to this constraint.
* `dtplyr` is missing timings for functions that are not yet implemented in the package.
* `pandas` comparisons are in the process of being added - more will be added soon.
* All tests are run 5 times. The times shown are the median of those 10 runs.
* All timings are in milliseconds.
* All tests can be found in the source code [here](https://github.com/markfairbanks/tidytable/blob/master/vignettes/speed_comparisons.Rmd).
* FAQ - Why are some `tidytable` functions faster than their `data.table` counterpart?
  + All R functions have some slight natural variation in their execution time. 
  If a `tidytable` function appears to be "faster" than `data.table` it's due to this.
  Overall these timings are to highlight that the "time cost" of translating `tidyverse` syntax
  to `data.table` syntax is negligible to the user.
* Lastly I'd like to mention that these tests were not rigorously created to cover all angles equally. They are just meant to be used as general insight into the performance of these packages.

```{r benchmark_setup, echo=FALSE}
pacman::p_load(tidytable, tidyverse, data.table, dtplyr)
setDTthreads(4)

iters <- 5

set.seed(123)
initial_data_size <- 10000000
data_size <- initial_data_size
test_dt <- tidytable(a = sample(1:20, data_size, TRUE),
                     b = sample(1:20, data_size, TRUE),
                     c = sample(c("a","a","b","c","d"), data_size, TRUE),
                     d = sample(c("a","a","b","c","d"), data_size, TRUE))

test_tbl <- as_tibble(test_dt)
```

```{r filter, echo=FALSE}
inv_gc(); inv_gc();

filter_marks <- bench::mark(
  tidyverse = filter(test_tbl, a <= 7, c == "a"),
  dtplyr = collect(filter(lazy_dt(test_dt), a <= 7, c == "a")),
  tidytable = filter.(test_dt, a <= 7, c == "a"),
  data.table = test_dt[a <= 7 & c == "a"],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "filter")

inv_gc(); inv_gc();
```

```{r summarize, echo=FALSE}
summarize_marks <- bench::mark(
  tidyverse = summarize(group_by(test_tbl, c), avg_a = mean(a), .groups = "drop"),
  dtplyr = collect(summarize(group_by(lazy_dt(test_dt), c), avg_a = mean(a))),
  tidytable = summarize.(test_dt, avg_a = mean(a), .by = c),
  data.table = test_dt[, list(avg_a = mean(a)), by = c],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "summarize")

inv_gc(); inv_gc();
```

```{r mutate, echo=FALSE}
mut_dt <- copy(test_dt)

mutate_marks <- bench::mark(
  tidyverse = mutate(test_tbl, double_a = a * 2, a_plus_b = a + b),
  dtplyr = collect(mutate(lazy_dt(test_dt), double_a = a * 2, a_plus_b = a + b)),
  tidytable = mutate.(test_dt, double_a = a * 2, a_plus_b = a + b),
  data.table = mut_dt[, ':='(double_a = a * 2, a_b = a + b)][],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "mutate")

inv_gc(); inv_gc();
```

```{r arrange, echo=FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$")); inv_gc(); inv_gc();

data_size <- initial_data_size * .1
test_dt <- tidytable(a = sample(1:20, data_size, TRUE),
                     b = sample(1:20, data_size, TRUE),
                     c = sample(c("a","a","b","c","d"), data_size, TRUE),
                     d = sample(c("a","a","b","c","d"), data_size, TRUE))

test_tbl <- as_tibble(test_dt)

arrange_marks <- bench::mark(
  tidyverse = arrange(test_tbl, c, a),
  dtplyr = collect(arrange(lazy_dt(test_dt), c, a)),
  tidytable = arrange.(test_dt, c, a),
  data.table = test_dt[order(c, a)],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "arrange")
```

```{r distinct, echo=FALSE}
distinct_marks <- bench::mark(
  tidyverse = distinct(test_tbl),
  dtplyr = collect(distinct(lazy_dt(test_dt))),
  tidytable = distinct.(test_dt),
  data.table = unique(test_dt),
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "distinct")

inv_gc(); inv_gc();
```

```{r fill, echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$")); inv_gc(); inv_gc();

size <- initial_data_size * .2
fill_dt <- tidytable(
  id = sample(1:3, size, replace = TRUE),
  # chr = sample(c("a","a","a","b","b",NA), size, replace = TRUE),
  int1 = sample(c(1:5,NA), size, replace = TRUE),
  int2 = sample(c(1:5,NA), size, replace = TRUE))

fill_tbl <- as_tibble(fill_dt)

fill_marks <- bench::mark(
  tidyverse = fill(group_by(fill_tbl, id), int1, int2),
  tidytable = fill.(fill_dt, int1, int2, .by = id),
  # needs shallow copy or subsequent runs won't have NAs to fill
  data.table = tidytable:::shallow(fill_dt)[, c("int1", "int2") := lapply(.SD, nafill, type = "locf"),
                                            by = id],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "fill")
```

```{r case, echo=FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$")); inv_gc(); inv_gc();

data_size <- initial_data_size * .3
x <- rnorm(data_size)
median_x <-  median(x)

case_marks <- bench::mark(
  tidyverse = case_when(x < median_x ~ 1,
                        x >= median_x ~ 2,
                        TRUE ~ 3),
  tidytable = case.(x < median_x, 1,
                    x >= median_x, 2,
                    default = 3),
  data.table = fcase(x < median_x, 1,
                     x >= median_x, 2,
                     rep(TRUE, length(x)), 3,
                     default = NA),
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "case_when")
```

```{r pivot_wider, echo=FALSE}
remove(x, median_x)

remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$")); inv_gc(); inv_gc();

data_size <- initial_data_size/5000
wider_tbl <- expand_grid(case = as.character(seq(1, data_size)),
                         year = seq(1900, data_size),
                         name = c("x", "y", "z")) %>%
  mutate(value = rnorm(nrow(.)))

wider_dt <- as_tidytable(wider_tbl)

wider_marks <- bench::mark(
  tidyverse = pivot_wider(wider_tbl, names_from = name, values_from = value),
  tidytable = pivot_wider.(wider_dt, names_from = name, values_from = value),
  data.table = data.table::dcast.data.table(wider_dt, case + year ~ name),
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "pivot_wider")
```

```{r pivot_longer, echo=FALSE}
longer_dt <- pivot_wider.(wider_dt, names_from = name, values_from = value)

longer_tbl <- as_tibble(longer_dt)

remove(wider_dt, wider_tbl); inv_gc(); inv_gc();

longer_marks <- bench::mark(
  tidyverse = pivot_longer(longer_tbl, cols = c(x, y, z)),
  tidytable = pivot_longer.(longer_dt, cols = c(x, y, z)),
  data.table = data.table::melt(longer_dt, measure.vars = c("x","y","z"),
                                variable.name = "name", variable.factor = FALSE),
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "pivot_longer")
```

```{r left_join, echo=FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$")); inv_gc(); inv_gc();

data_size <- initial_data_size * .1
left_dt <- tidytable(a = sample(1:20, data_size, TRUE),
                      b = sample(1:20, data_size, TRUE),
                      c = sample(c("a","a","b","c","d"), data_size, TRUE),
                      d = sample(c("a","a","b","c","d"), data_size, TRUE))

right_dt <- tidytable(c = c("a","b","c","d"),
                      e = c(1,2,3,4))

left_tbl <- as_tibble(left_dt)
right_tbl <- as_tibble(right_dt)

left_join_marks <- bench::mark(
  tidyverse = left_join(left_tbl, right_tbl, by = "c"),
  dtplyr = collect(left_join(lazy_dt(left_tbl), lazy_dt(right_tbl), by = "c")),
  tidytable = left_join.(left_dt, right_dt),
  data.table = right_dt[left_dt, on = "c", allow.cartesian = TRUE],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "left_join")
```

```{r inner_join, echo=FALSE}
inner_join_marks <- bench::mark(
  tidyverse = inner_join(left_tbl, right_tbl, by = "c"),
  dtplyr = collect(inner_join(lazy_dt(left_tbl), lazy_dt(right_tbl), by = "c")),
  tidytable = inner_join.(left_dt, right_dt),
  data.table = left_dt[right_dt, on = "c", allow.cartesian = TRUE, nomatch = 0],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "inner_join")
```

```{r unnest, echo = FALSE}
data_size <- initial_data_size * .03
list_dt <- data.table(x = sample(1:20, data_size, TRUE))

unnest_dt <- data.table(id = 1:5,
                        list_column = list(list_dt, list_dt, list_dt, list_dt, list_dt))

unnest_tbl <- unnest_dt %>%
  mutate(list_column = map(list_column, as_tibble))

unnest_marks <- bench::mark(
  tidyverse = tidyr::unnest(unnest_tbl, list_column),
  tidytable = unnest.(unnest_dt, list_column),
  data.table = unnest_dt[, unlist(list_column, recursive = FALSE), by = id],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  dplyr::mutate(expression = as.character(expression),
                func_tested = "unnest")
```

```{r nest, echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$")); inv_gc(); inv_gc();

data_size <- initial_data_size * .03
list_df <- data.table(x = sample(1:20, data_size, TRUE))

nest_dt <- data.table(id = 1:5,
                      list_column = list(list_df, list_df, list_df, list_df, list_df)) %>%
  unnest.(list_column)

nest_tbl <- as_tibble(nest_dt)

nest_marks <- bench::mark(
  tidyverse = group_nest(nest_tbl, id),
  tidytable = nest_by.(nest_dt, id),
  data.table = nest_dt[, list(data = list(.SD)), by = id],
  check = FALSE, iterations = iters, memory = FALSE, filter_gc = FALSE, time_unit = 'ms') %>%
  mutate(expression = as.character(expression),
         func_tested = "nest")
```

```{r echo = FALSE}
# # pandas code
# import pandas as pd
# import numpy as np
# import gc
# data_size = 10000000
#
# test_df = pd.DataFrame(
#     {'a': np.random.choice(np.arange(20), data_size),
#      'b': np.random.choice(np.arange(20), data_size),
#      'c': np.random.choice(['a','a','b','c','d'], data_size),
#      'd': np.random.choice(['a','a','b','c','d'], data_size)}).convert_dtypes()
#
# print('filter')
# %timeit test_df.loc[(test_df.a <= 7) & (test_df.c == 'a')]
# print('summarize')
# %timeit test_df.groupby('c')['a'].mean()
# print('mutate pt1')
# %timeit test_df['double_a'] = test_df.a * 2
# print('mutate pt2')
# %timeit test_df['a_plus_b'] = test_df.a + test_df.b
#
# del test_df
# gc.collect()
#
# arrange_size = 1000000
#
# arrange_df = pd.DataFrame(
#     {'a': np.random.choice(np.arange(20), arrange_size),
#      'b': np.random.choice(np.arange(20), arrange_size),
#      'c': np.random.choice(['a','a','b','c','d'], arrange_size),
#      'd': np.random.choice(['a','a','b','c','d'], arrange_size)}).convert_dtypes()
#
# print('arrange')
# %timeit arrange_df.sort_values(['c', 'a'])
# print('distinct')
# %timeit arrange_df.drop_duplicates()
# del arrange_df
# gc.collect()
#
# case_df = pd.DataFrame({'x': np.random.normal(size = 3000000)})
#
# case_df['median_x'] = case_df['x'].median()
#
# print('case_when')
# %timeit case_df['test'] = np.where(case_df.x < case_df.median_x, case_df.x * 2, np.where(case_df.x >= case_df.median_x, case_df.x * 3, case_df.x))
#
# del case_df
# gc.collect()
#
# fill_size = 2000000
# fill_df = pd.DataFrame({'x': np.arange(fill_size),
#                         'id': np.random.choice(np.arange(3), fill_size)})
#
# fill_df = fill_df.assign(y = fill_df.x.shift(10), z = fill_df.x.shift(-10)).convert_dtypes()
#
# print('fill')
# %timeit fill_df.groupby('id')[['x','y']].fillna(method = 'ffill')
```

```{r all_marks, echo = FALSE, message = FALSE, paged.print = FALSE}
library(magrittr)
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$")); inv_gc(); inv_gc();

all_marks <- ls(pattern="marks$") %>%
  lapply(get) %>%
  map(~.x %>%
        select(func_tested, expression, median)
      ) %>%
  bind_rows()

all_marks <- all_marks %>%
  mutate_if(is.character, str_trim) %>%
  mutate(median = round(median, 1)) %>%
  pivot_wider.(names_from = expression, values_from = median) %>%
  as_tibble() %>%
  mutate(tidytable_vs_dplyr = scales::percent(tidytable/tidyverse, accuracy = .1)) %>%
  # mutate_at(vars(tidyverse, tidytable, data.table, dtplyr), ~str_c(.x, "ms")) %>%
  arrange(func_tested)
```

```{r pandas, echo = FALSE, message = FALSE}
pandas_marks <- tibble(arrange = 355,
                       case_when = 59.2,
                       distinct = 309,
                       filter = 707,
                       fill = 846,
                       mutate = 86.4,
                       summarize = 834) %>%
  pivot_longer(everything(), names_to = "func_tested", values_to = "pandas")

all_marks <- all_marks %>%
  left_join(pandas_marks) %>%
  select.(func_tested, data.table, tidytable, dtplyr, tidyverse, pandas, tidytable_vs_dplyr)
```

```{r echo=FALSE}
all_marks
```
