---
title: "speed_comparisons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{speed_comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  paged.print = FALSE
)
```

### Speed Comparisons

A few notes:

* Comparing times from separate functions won't be very useful. For example - the `summarize()` tests were performed on a different dataset from `case_when()`.
* `setDTthreads(4)` was used for `data.table` & `tidytable` timings.
* Modify-by-reference was used in `data.table` when being compared to `mutate.()` & `dplyr::mutate()`
* `fill.()` & `tidyr::fill()` both work with character/factor/logical columns, whereas `data.table::nafill()`
does not. Testing only included numeric columns due to this constraint.
* `dtplyr` is missing timings for functions that are not yet implemented in the package.
* `pandas` comparisons are in the process of being added - more will be added soon.
* All tests are run 11 times. The times shown are the median of those 11 runs.
* All timings are in milliseconds.
* All tests can be found in the source code [here](https://github.com/markfairbanks/tidytable/blob/master/vignettes/speed_comparisons.Rmd).
* FAQ - Why are some `tidytable` functions faster than their `data.table` counterpart?
  + Short answer - they're not! After all they're just using `data.table` in the background.
  + Long answer - All R functions have some slight natural variation in their execution time. 
  By chance the `tidytable` runs were slightly shorter on those specific functions on this
  iteration of the tests. However one goal of these tests is to show that the "time cost" of
  translating `tidyverse` syntax to `data.table` is very negligible to the
  user (especially on medium-to-large datasets).
* Lastly I'd like to mention that these tests were not rigorously created to cover all angles equally. They are just meant to be used as general insight into the performance of these packages.

```{r echo = FALSE}
pacman::p_load(tidytable, dtplyr, data.table)

setDTthreads(4)

set.seed(123)

random_strings <- c('OoVt', 'wCbu', 'cXxX', 'jdFu', 'MCRx',
                    'ukhz', 'ikce', 'PHyu', 'jpBY', 'nLQM')

bench_mark <- function(..., .fn = NULL) {
  bench::mark(...,
              check = FALSE, iterations = 11,
              memory = FALSE, time_unit = 'ms') %>%
    suppressWarnings() %>%
    mutate(expression = as.character(expression),
            median = round(median, 1),
            func_tested = .env$.fn) %>%
    select(func_tested, expression, median) %>%
    pivot_wider(names_from = expression, values_from = median)
}

test_data <- function(.size) {
  tidytable(a = sample(1:20, .size, TRUE),
            b = sample(1:20, .size, TRUE),
            c = sample(random_strings, .size, TRUE),
            d = sample(random_strings, .size, TRUE))
}

data_size <- 10000000

test_dt <- test_data(data_size)

test_tbl <- dplyr::as_tibble(test_dt)
```

```{r echo = FALSE}
inv_gc(); inv_gc();

filter_marks <- bench_mark(
  tidyverse = dplyr::filter(test_tbl, a <= 7, c == 'OoVt'),
  dtplyr = as.data.table(dplyr::filter(lazy_dt(test_dt), a <= 7, c == 'OoVt')),
  tidytable = filter(test_dt, a <= 7, c == 'OoVt'),
  data.table = test_dt[a <= 7 & c == 'OoVt'],
  .fn = "filter"
)

inv_gc(); inv_gc();

summarize_marks <- bench_mark(
  tidyverse = test_tbl %>%
    dplyr::group_by(c, d) %>%
    dplyr::summarize(avg_a = mean(a), med_b = median(b), .groups = "drop"),
  dtplyr = lazy_dt(test_dt) %>%
    dplyr::group_by(c, d) %>%
    dplyr::summarize(avg_a = mean(a), med_b = median(b), .groups = "drop") %>%
    as.data.table(),
  tidytable = summarize(test_dt, avg_a = mean(a), med_b = median(b), .by = c(c, d)),
  data.table = test_dt[, list(avg_a = mean(a), med_b = median(b)), by = .(c, d)],
  .fn = "summarize"
)

inv_gc(); inv_gc();

mut_dt <- copy(test_dt)

mutate_marks <- bench_mark(
  tidyverse = dplyr::mutate(test_tbl, double_a = a * 2, a_plus_b = a + b),
  dtplyr = as.data.table(dplyr::mutate(lazy_dt(test_dt), double_a = a * 2, a_plus_b = a + b)),
  tidytable = mutate(test_dt, double_a = a * 2, a_plus_b = a + b),
  data.table = mut_dt[, ':='(double_a = a * 2, a_b = a + b)][],
  .fn = "mutate"
)
```

```{r echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

test_dt <- test_data(data_size * .1)

test_tbl <- dplyr::as_tibble(test_dt)

arrange_marks <- bench_mark(
  tidyverse = dplyr::arrange(test_tbl, c, a),
  dtplyr = as.data.table(dplyr::arrange(lazy_dt(test_dt), c, a)),
  tidytable = arrange(test_dt, c, a),
  data.table = test_dt[order(c, a)],
  .fn = "arrange"
)

inv_gc(); inv_gc();

distinct_marks <- bench_mark(
  tidyverse = dplyr::distinct(test_tbl),
  dtplyr = as.data.table(dplyr::distinct(lazy_dt(test_dt))),
  tidytable = distinct(test_dt),
  data.table = unique(test_dt),
  .fn = "distinct"
)
```

```{r echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

fill_size <- data_size * .2
fill_dt <- tidytable(
  id = sample(1:3, fill_size, replace = TRUE),
  int1 = sample(c(1:5, NA), fill_size, replace = TRUE),
  int2 = sample(c(1:5, NA), fill_size, replace = TRUE)
)

fill_tbl <- dplyr::as_tibble(fill_dt)

fill_marks <- bench_mark(
  tidyverse = tidyr::fill(group_by(fill_tbl, id), int1, int2),
  dtplyr = as.data.table(tidyr::fill(dplyr::group_by(lazy_dt(fill_dt), id), int1, int2)),
  tidytable = fill(fill_dt, int1, int2, .by = id),
  # needs a copy or subsequent runs won't have NAs to fill
  data.table = tidytable:::shallow(fill_dt)[
    ,
    c("new1", "new2") := lapply(.SD, nafill, type = "locf"),
    .SDcols = c("int1", "int2"),
    by = id],
  .fn = "fill"
)
```

```{r echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

case_size <- data_size * .3
x <- runif(case_size * .3)

case_marks <- bench_mark(
  tidyverse = dplyr::case_when(x < .5 ~ 1,
                        x >= .5 ~ 2,
                        TRUE ~ 3),
  tidytable = case(x < .5, 1,
                   x >= .5, 2,
                   default = 3),
  data.table = fcase(x < .5, 1,
                     x >= .5, 2,
                     rep(TRUE, length(x)), 3),
  .fn = "case_when"
)
```

```{r echo = FALSE}
remove(x)

remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

wider_dt <- 1:3 %>%
  map_dfr(
    ~ expand_grid(a = letters, b = LETTERS, c = paste0(letters, LETTERS)) %>%
    unite("name") %>%
    mutate(value = sample(as.double(1:n()), n())) %>%
    mutate(id = .x)
  ) %>%
  relocate(id)

wider_tbl <- dplyr::as_tibble(wider_dt)

wider_marks <- bench_mark(
  tidyverse = tidyr::pivot_wider(wider_tbl, names_from = name, values_from = value),
  dtplyr = lazy_dt(wider_dt) %>%
    tidyr::pivot_wider(names_from = name, values_from = value) %>%
    as.data.table(),
  tidytable = pivot_wider(wider_dt, names_from = name, values_from = value),
  data.table = data.table::dcast.data.table(wider_dt, id ~ name),
  .fn = "pivot_wider"
)
```

```{r echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

longer_size <- 100000
longer_dt <-
  tidytable(id = 1:longer_size) %>%
  bind_cols(
    map_dfc(letters, ~ tidytable(!!.x := sample(c(letters, LETTERS), longer_size, TRUE)))
  )

longer_tbl <- dplyr::as_tibble(longer_dt)

longer_marks <- bench_mark(
  tidyverse = tidyr::pivot_longer(longer_tbl, cols = -id),
  dtplyr = as.data.table(tidyr::pivot_longer(lazy_dt(longer_dt), cols = -id)),
  tidytable = pivot_longer(longer_dt, cols = -id, fast_pivot = TRUE),
  data.table = melt(longer_dt,
                    measure.vars = names(longer_dt)[names(longer_dt) != "id"],
                    variable.name = "name",
                    variable.factor = FALSE),
  .fn = "pivot_longer"
)
```

```{r echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

left_dt <- test_data(data_size * .1)

right_dt <- left_dt %>%
  distinct(c, d) %>%
  mutate(e = row_number())

left_tbl <- dplyr::as_tibble(left_dt)
right_tbl <- dplyr::as_tibble(right_dt)

left_join_marks <- bench_mark(
  tidyverse = dplyr::left_join(left_tbl, right_tbl, by = c("c", "d")),
  dtplyr = as.data.table(dplyr::left_join(lazy_dt(left_dt), lazy_dt(right_dt), by = c("c", "d"))),
  tidytable = left_join(left_dt, right_dt, by = c("c", "d")),
  data.table = right_dt[left_dt, on = c("c", "d"), allow.cartesian = TRUE],
  .fn = "left_join"
)

inv_gc(); inv_gc();

inner_join_marks <- bench_mark(
  tidyverse = dplyr::inner_join(left_tbl, right_tbl, by = c("c", "d")),
  dtplyr = as.data.table(dplyr::inner_join(lazy_dt(left_dt), lazy_dt(right_dt), by = c("c", "d"))),
  tidytable = inner_join(left_dt, right_dt, c("c", "d")),
  data.table = left_dt[right_dt, on = c("c", "d"), allow.cartesian = TRUE, nomatch = 0],
  .fn = "inner_join"
)
```

```{r echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

unnest_dt <- test_data(data_size * .3) %>%
  nest_by(c, d, .key = "list_column")

unnest_tbl <- unnest_dt %>%
  mutate(list_column = map(list_column, dplyr::as_tibble)) %>%
  dplyr::as_tibble()

unnest_marks <- bench_mark(
  tidyverse = tidyr::unnest(unnest_tbl, list_column),
  tidytable = unnest(unnest_dt, list_column),
  data.table = unnest_dt[, unlist(list_column, recursive = FALSE), by = .(c, d)],
  .fn = "unnest"
)
```

```{r echo = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

nest_dt <- test_data(data_size * .1)

nest_tbl <- dplyr::as_tibble(nest_dt)

nest_marks <- bench_mark(
  tidyverse = dplyr::group_nest(nest_tbl, c, d),
  dtplyr = as.data.table(tidyr::nest(lazy_dt(nest_dt), data = c(c, d))),
  tidytable = nest_by(nest_dt, c, d),
  data.table = nest_dt[, list(data = list(.SD)), by = .(c, d)],
  .fn = "nest"
)
```

```{r echo = FALSE}
# # pandas code (run in a .ipynb in vscode)
# import pandas as pd
# import numpy as np
# import gc
# from timeit import timeit
# 
# num_runs = 5
# def run_benchmarks(_dict):
#     return {key: np.round(timeit(value, number = num_runs) * 1000/num_runs, 1) for key, value in _dict.items()}
# 
# np.random.seed(123)
# random_strings = np.array(['OoVt', 'wCbu', 'cXxX', 'jdFu', 'MCRx', 'ukhz', 'ikce', 'PHyu', 'jpBY', 'nLQM'])
# 
# initial_data_size = 10000000
# data_size = initial_data_size
# 
# test_df = (
#     pd.DataFrame({
#         'a': np.random.choice(np.arange(20), data_size),
#         'b': np.random.choice(np.arange(20), data_size),
#         'c': np.random.choice(random_strings, data_size),
#         'd': np.random.choice(random_strings, data_size)
#     })
#     .convert_dtypes()
# )
# 
# bench_fns1 = dict(
#     filter = lambda: test_df.loc[(test_df.a <= 7) & (test_df.c == 'OoVt')],
#     summarize = lambda: test_df.groupby(['c', 'd']).agg(avg_a = ('a', np.mean), med_b = ('b', np.median)),
#     mutate = lambda: test_df.assign(double_a = test_df.a * 2, a_plus_b = test_df.a + test_df.b)
# )
# 
# bench_marks1 = run_benchmarks(bench_fns1)
# 
# data_size = initial_data_size * .1
# test_df = test_df.head(int(data_size))
# gc.collect()
# 
# bench_fns2 = dict(
#     arrange = lambda: test_df.sort_values(['c', 'a']),
#     distinct = lambda: test_df.drop_duplicates()
# )
# 
# bench_marks2 = run_benchmarks(bench_fns2)
# 
# del test_df
# gc.collect()
# 
# x = np.random.normal(size = 3000000)
# median_x = np.median(x)
# 
# bench_fns3 = dict(
#     case_when = lambda: np.select(
#         [x < median_x, x <= median_x],
#         [1, 2],
#         default = 3
#     )
# )
# 
# bench_marks3 = run_benchmarks(bench_fns3)
# 
# del x, median_x
# gc.collect()
# 
# data_size = int(initial_data_size * .2)
# fill_df = (
#     pd.DataFrame(dict(
#         id = np.random.choice(np.arange(3), data_size),
#         int1 = np.random.choice(np.array([1, 2, 3, 4, 5, np.nan]), data_size),
#         int2 = np.random.choice(np.array([1, 2, 3, 4, 5, np.nan]), data_size)
#     ))
#     .convert_dtypes()
# )
# 
# bench_fns4 = dict(
#     fill = lambda: fill_df.groupby('id')[['int1','int2']].fillna(method = 'ffill')
# )
# 
# bench_marks4 = run_benchmarks(bench_fns4)
# 
# del fill_df
# gc.collect()
# 
# bench_marks = bench_marks1 | bench_marks2 | bench_marks3 | bench_marks4
# 
# dict(sorted(bench_marks.items()))
```

```{r echo = FALSE, message = FALSE}
remove(list = ls(pattern = "tbl$")); remove(list = ls(pattern = "dt$"));
inv_gc(); inv_gc();

marks_names <- setdiff(
  unlist(ls(pattern="marks$")),
  c("all_marks", "pandas_marks")
)

all_marks <- marks_names %>%
  map_dfr(get) %>%
  arrange(func_tested)
```

```{r pandas, echo = FALSE, message = FALSE}
pandas_marks <-
  tidytable(
    arrange = 715.8,
    case_when = 64.4,
    distinct = 308.7,
    fill = 724,
    filter = 903.8,
    mutate = 780.1,
    summarize = 3079.8
  ) %>%
  pivot_longer(everything(), names_to = "func_tested", values_to = "pandas")

all_marks <- all_marks %>%
  left_join(pandas_marks) %>%
  select(func_tested, data.table, tidytable, dtplyr, tidyverse, pandas)
```

```{r echo = FALSE}
print(glue::glue("Date last run: {lubridate::today()}"))
```

```{r echo = FALSE}
all_marks
```
